The impact of earthquake cycle variability on neotectonic and paleoseismic slip rate estimates

### 2018-12-10 manuscript/eq_cycle_slip_rates.tex
finished revised manuscript

\documentclass[se, manuscript]{copernicus}
\usepackage{mathptmx}

@@

\begin{abstract}
Because of the natural (aleatoric) variability in earthquake
recurrence intervals and coseismic displacements on a fault, cumulative
slip on a fault does not increase linearly or perfectly step-wise with
time; instead, some amount of variability in shorter-term slip rates
results. Though this variability could greatly affect the accuracy of
neotectonic (i.e., late Quaternary) and paleoseismic slip rate
estimates, these effects have not been quantified. In this study,
idealized faults with four different, representative earthquake
recurrence distributions are created with equal mean recurrence
intervals (1,000 years) and coseismic slip distributions, and the
variability in slip rate measurements over 500 to 100,000 year
measurement windows is calculated for all faults through Monte Carlo
simulations. The recurrence distributions used are quasi-periodic,
unclustered and clustered lognormal distributions, and an unclustered
exponential distribution. The results demonstrate that the most
important parameter is the coefficient of variation (\(CV\) = standard
deviation / mean) of the recurrence distributions rather than the shape
of the distribution itself. Slip rate variability over short time scales
(\textless{} 5,000 years or 5 mean earthquake cycles) is quite high, varying by 
    a factor of 3 or more from the mean,
but decreases with time and is close to stable after
\textasciitilde{}40,000 years (40 mean earthquake cycles). This
variability is higher for recurrence distributions with a higher
\(CV\). The natural variability in the slip rate estimates compared to
the true value is then used to estimate the epistemic uncertainty in a
single slip rate measurement (as one would make in a geological study)
in the absence of any measurement uncertainty. This epistemic
uncertainty is very high (a factor of 2 or more) for measurement windows
of a few mean earthquake cycles (as in a paleoseismic slip rate
estimate), but decreases rapidly to a factor of 1—2 with \textgreater{}5
mean earthquake cycles (as in a neotectonic slip rate study). These
uncertainties are independent of, and should be propagated with
uncertainties in fault displacement and geochronologic measurements used
to estimate slip rates. They may then aid in the comparison of slip
rates from different methods or the evaluation of potential slip rate
changes over time.
\end{abstract}

@@

Fault slip rates are generally estimated by dividing measurements of the
offset of geologic marker features by the time over which that offset
accumulated. The uncertainty in the resulting slip estimate is typically
treated as \emph{epistemic}, and quantified through the propagation of
the measurement uncertainties on the offset and time quantities
\citep[e.g.,][]{bird_uncertainties_2007,zechar_incorporating_2009}.
However, for slip rate estimates on active faults made from offset
measurements near the fault trace (i.e., within a horizontal distance
that is a small fraction of the fault's locking depth, as the width of the zone 
affected by earthquake-cycle strains is a function of locking depth 
\citep[e.g.,][]{savage_geodetic_1973, hetland_interseismic_2006}), the episodic
nature of surface displacement due to the earthquake cycle will
necessarily affect the results: If the measurements are taken
immediately before an earthquake, the measured offset and resulting slip
rate estimate will be lower than average, while if the measurements are
taken immediately after an earthquake, the offset and rate will be
higher.

The magnitude of the perturbation to the slip rate is, of course, a
function of the number of cumulative earthquakes that have contributed
to the measured offset (plus any aseismic strain such as afterslip or creep).
For older Quaternary markers that have experienced tens to hundreds of major
earthquakes, the effects will be minor, and for bedrock geologic markers with
kilometers of displacement, the earthquake cycle is likely not worth accounting
for. However, due to progressive erosion of geologic markers and the challenge
of dating many late Pliocene to early Quaternary units (which are too old for
radiocarbon and many cosmogenic nuclide systems), geologists often have no
choice but to choose late Pleistocene to Holocene markers to date. These
units may also be more desirable targets if the scientists are primarily
concerned with estimating the contemporary slip rate on a fault with a slip
rate that may vary over Quaternary timescales
\citep[e.g.,][]{rittase_temporal_2014,zinke_highly_2018}. For slow-moving 
faults, the slip either long waiting to be released, or recently released, may 
represent a sizeable fraction of the measured fault offset.

@@

The most commonly used is the \emph{exponential} distribution. This is
associated with a Poisson process, and is the distribution that results
from earthquakes being distributed uniformly randomly within some time interval.
Consequently, the probability of an earthquake (or other event) occurring at any
time does not change with time since the previous event (in other words, the
hazard function is time-invariant); this leads to the characterization of the
exponential recurrence distribution as `random', `memoryless' or
`time-independent'. The exponential distribution is also the simplest to
describe statistically, as it requires only one parameter (the mean rate
parameter), which is the inverse of the statistical scale parameter (the scale 
parameter of a distribution determines the dispersion of the distribution, 
while the shape parameter determines the shape or form of the distribution; a 
parameter that shifts the distribution along the $x$-axis is called a location 
parameter).  The standard deviation of a large number of samples generated from 
an exponential distribution is equal to the mean.

@@

Clustered earthquake sequences have sets of very tightly spaced
earthquakes that are widely separated, and have a \(CV>1\). These may
be generated from a hyperexponential distribution, which is the sum of
multiple exponentials with different means, or from the time-dependent
distributions given above, given the right parameters.

No consensus exists among earthquake scientists as to the most
appropriate recurrence interval distribution. As is generally the case
with propriety, the safest and probably most correct assumption is that
it is context-dependent. Many studies of plate boundary faults such as
the San Andreas conclude that major or `characteristic' earthquakes are
periodic\\
\citep[e.g.,][]{berryman_major_2012, scharer_quasi-periodic_2010}.
Conversely, many intraplate faults with low slip rates appear to show
clustered earthquakes separated by long intervals of seismic quiescence
\citep[e.g.,][]{clark_long-term_2012}. However, one can find examples of
studies indicating the opposite conclusions
\citep{tuttle_earthquake_2002, grant_paleoseismic_1994}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  A \emph{periodic} distribution, represented by a lognormal
  distribution with a mean recurrence interval \(\mu\) = 1000 years, and
  a standard deviation \(\sigma\) = 500 years, and a \(CV\) =
  0.5 \footnote{Please note that $\mu$ and $\sigma$ do not represent the scale
  and shape parameters of the lognormal distribution, though these symbols are
  commonly used to represent these parameters, as the scale and shape parameters
  of a lognormal distribution $L$ are the standard deviation and mean of the
  normal distribution $\ln L$. Instead, the shape parameter $sh$
  and scale parameter $sc$ are derived from $\mu$ and $\sigma$ through the
  relations $sh = \sqrt{1 + \sigma^2 / \mu^2}$ and 
  $sc = \ln [ \mu / \sqrt{1 + \sigma^2 / \mu^2}]$.}
\item
  An \emph{unclustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 1000 years, and a \(CV\) =
  1.0.
\item
  An \emph{clustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 2000 years, and a \(CV\) =
  2.0.
\item
  An \emph{unclustered} time-independent distribution, represented by an
  exponential distribution with a mean recurrence interval \(\mu\) =
  1000 years, a standard deviation \(\sigma\) = 1000 years, and a
  \(CV\) = 1.0.
\end{enumerate}

@@

All earthquake recurrence distributions share a single earthquake slip
distribution (Figure \ref{slip_dist}). This distribution is a lognormal
distribution with \(\mu\) = 1 m and \(\sigma\) = 0.75 m, which produces
essentially `characteristic' earthquakes that still nonetheless have
some variability. This is representative of behavior observed in many
studies \citep[e.g.,][]{zielke_slip_2010, klinger_characteristic_2011,
zielke_earthquake_2018}. Taken together, the mean slip of 1 m and the
mean recurrence interval of 1000 years shared by each of the recurrence
interval distributions yields a mean slip rate of 1 mm/yr. This rate is
fairly typical for intraplate faults studied by paleoseismologists, and
also allows for easy normalization so that the results of this study can
be generalized to faults with different parameters.

@@

The construction of the displacement histories is straightforward. From
each recurrence distribution, a little over 2,000 samples are drawn
randomly. Then, these are combined with an equal number of displacement
samples drawn randomly from the earthquake slip distribution. Finally a
cumulative displacement history is created for each series from a
cumulative sum of both the recurrence interval samples (producing an
earthquake time series) and displacement samples (producing a cumulative
slip history). Years with no earthquakes are represented as having no
increase in cumulative displacement. Then, the series is trimmed at
year=2,000,000; it is initially made longer because the stochastic
nature of the sample sets means that 2,000 earthquakes may not always
reach 2,000,000 years.

@@

Please note that in the construction of the cumulative displacement
histories, all samples are independent. This means that the duration of any
recurrence interval does not depend on the duration of the previous or
subsequent interval (in other words, there is no autocorrelation in these
series); the same applies to the displacement samples. Furthermore, the
magnitude of displacement is independent of the corresponding recurrence
interval. It is currently unknown to what degree autocorrelation exists in
real earthquake time and displacement series, or how much correlation is
present between recurrence intervals and subsequent displacements.
Autocorrelation in recurrence interval sequences is essentially unstudied,
though on the basis of preliminary, unreviewed analysis
\citep{styron_survival_2017}, I suspect that it is as important as \(CV\).
With respect to the amount of slip per event and loading time, the framework
of elastic rebound theory in its most basic form should predict some
correspondence between inter-event (loading) duration and slip magnitude, and
this is included (implicitly or explicitly) in oscillator models
incorporating complete stress or strain release in each earthquake
\citep[e.g.,][]{matthews_brownian_2002,dicaprio_post-seismic_2008} or in any
model where coseismic friction drops to zero, as this is functionally
equivalent (because \(f^a = \tau^a_s / \tau^a_n\), where \(f^a\),
\(\tau^a_s\) and \(\tau^a_n\) are respectively friction at rupture arrest,
shear stress at rupture arrest, and effective normal stress at rupture
arrest, zero friction implies zero shear stress, or complete stress drop).
Given a reasonably constant loading rate, complete shear stress or strain
release implies some proportionality between the loading time and
displacement. Nonetheless, this correspondence is not found in the more
extensive paleoseismic datasets, such as those by
\citet{benedetti_earthquake_2013} (or the correlation may be negative as
found by \citet{weldon_wrightwood_2004}), but the number of paleoseismic
datasets of sufficient size and cquality to identify these effects with
statistical significance are few indeed.

@@

A major goal of this study is to provide an answer to the question, \emph{How
long should slip rates be measured over in order to estimate a meaningful
rate?} This question will be answered by looking at the distribution in
\(\hat{R}\) as a function of \(t\). Fifty values of \(t\) from 500 years to
100,000 years, logarithmically spaced, are used. Note that given \(\mu\) of
1000 years, this translates to 0.5-100 mean numbers of earthquakes in the
window.

@@

Note that with a measurement time exceeding 5-10 mean earthquake cycles, the
standard error ($n / \sqrt{(\sigma_D^2 + \sigma_r^2)}$, where $\sigma_D$ is
the standard deviation of displacement, and $\sigma_r$ is the standard
deviation of the recurrence interval) is a reasonable approximation for the
standard deviation of the variability in the slip rates due to
earthquake-cycle stochasticity, and shows broadly similar decreasing
variation with increasing earthquake cycles. However, the standard error is
symmetrical, though the variability displayed here is asymmetrical due to the
asymmetry of the recurrence interval and displacement distributions.

@@

Second, the distributions are asymmetrical, especially the 5-95\%
interval. The 95th percentile is generally several times as far from the
measured value as the 5th percentile, meaning that the true value of the
slip rate may be much greater than the measured rate but not a
commensurately small fraction of the measured rate.

Third, the median rate before convergence at 5 \(\mu\) is greater than
the true rate, meaning that in most cases, very short-term slip rate
measurements will underestimate the true slip rate; this is a systematic
bias. This is a particular concern for paleoseismological slip rate
estimates, where there are rarely more than 5 events in any given
trench, though this bias will be decreased if slip rate measurements are made
from closed intervals only. However, the median is less than 2 times
\(\hat{R}\) following just 2-3 events, so the systematic bias is unlikely to
be much greater than the measurement uncertainties on the age or offset of
the events.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The variability in slip rates calculated over time windows less than 5
  mean earthquake cycles is very large, but begins to stabilize
  following \textasciitilde{}10-20 earthquakes.
\item
  The most important factor in controlling the variability in slip rate
  estimates is the coefficient of variation (\(CV\)); the different
  distributions themselves are relatively unimportant. Faults with
  \emph{periodic} earthquakes, with \(CV < 1\), have less initial
  variability, and stabilize rapidly. Faults with \emph{unclustered}
  earthquakes, with \(CV = 1\), have more variability and stabilize
  more slowly. Faults with \emph{clustered} earthquakes,\\
  with \(CV > 1\), have a great amount of initial variability and
  require a large number of earthquakes to stabilize.
\item
  The epistemic uncertainties around a measured slip rate are similarly
  large initially and then decrease with time. These uncertainties are
  initially biased, such that the measured slip rates typically
  underestimate the true slip rate for \textless{}5 mean earthquake
  cycles, but this fades with time. However, the uncertainties remain
  asymmetric, with a strong right skew.
\end{enumerate}

>>>>

\documentclass[se]{copernicus}
\usepackage{mathptmx}

@@

\begin{abstract}
Because of the natural (aleatoric) variability in earthquake recurrence
intervals and coseismic displacements on a fault, cumulative slip on a fault
does not increase linearly or perfectly step-wise with time; instead, some
amount of variability in shorter-term slip rates results. Though this
variability could greatly affect the accuracy of neotectonic (i.e., late
Quaternary) and paleoseismic slip rate estimates, these effects have not
been quantified. In this study, idealized faults with four different,
representative earthquake recurrence distributions are created with equal
mean recurrence intervals (1,000 years) and coseismic slip distributions,
and the variability in slip rate estimates over 500 to 100,000 year
measurement windows is calculated for all faults through Monte Carlo
simulations. Slip rates are calculated as net offset divided by elapsed
time, as in a typical neotectonic study. The recurrence distributions used
are quasi-periodic, unclustered and clustered lognormal distributions, and
an unclustered exponential distribution. The results demonstrate that the
most important parameter is the coefficient of variation (\(CV\) = standard
deviation / mean) of the recurrence distributions rather than the shape of
the distribution itself. Slip rate variability over short time scales
(\textless{}5,000 years or 5 mean earthquake cycles) is quite high, varying
by a factor of 3 or more from the mean, but decreases with time and is
close to stable after \textasciitilde{}40,000 years (40 mean earthquake
cycles). This variability is higher for recurrence distributions with a
higher \(CV\). The natural variability in the slip rate estimates compared
to the true value is then used to estimate the epistemic uncertainty in a
single slip rate measurement (as one would make in a geological study) in
the absence of any measurement uncertainty. This epistemic uncertainty is
very high (a factor of 2 or more) for measurement windows of a few mean
earthquake cycles (as in a paleoseismic slip rate estimate), but decreases
rapidly to a factor of 1—2 with \textgreater{}5 mean earthquake cycles (as
in a neotectonic slip rate study). These uncertainties are independent of,
and should be propagated with, uncertainties in fault displacement and
geochronologic measurements used to estimate slip rates. They may then aid
in the comparison of slip rates from different methods or the evaluation of
potential slip rate changes over time.
\end{abstract}

@@

Fault slip rates are generally estimated by dividing measurements of the offset
of geologic marker features by the time over which that offset accumulated (it
is not currently possible to measure a slip rate directly, though the term
`slip rate measurement' may be used to compare to a simulated or modeled
value). The uncertainty in the resulting slip estimate is typically treated as
\emph{epistemic}, and quantified through the propagation of the measurement
uncertainties on the offset and time quantities
\citep[e.g.,][]{bird_uncertainties_2007, zechar_incorporating_2009}.  However,
for slip rate estimates on active faults made from offset measurements near the
fault trace (i.e., within a horizontal distance that is a small fraction of the
fault's locking depth, as the width of the zone affected by earthquake-cycle
strains is a function of locking depth \citep[e.g.,][]{savage_geodetic_1973,
hetland_interseismic_2006}), the episodic nature of surface displacement due to
the fault's position in the earthquake cycle will necessarily affect the
results: If the measurements are taken immediately before an earthquake, the
measured offset and resulting slip rate estimate will be lower than average,
while if the measurements are taken immediately after an earthquake, the offset
and rate will be higher.

The magnitude of the perturbation to the slip rate estimate is, of course, a
function of the number of cumulative earthquakes that have contributed
to the measured offset (plus any aseismic strain such as afterslip or creep).
For older Quaternary markers that have experienced tens to hundreds of major
earthquakes, the effects will be minor, and for bedrock geologic markers with
kilometers of displacement, the earthquake cycle is likely not worth accounting
for. However, due to progressive erosion of geologic markers and the challenge
of dating many late Pliocene to early Quaternary units (which are too old for
radiocarbon and many cosmogenic nuclide systems), geologists often have no
choice but to choose late Pleistocene to Holocene markers to date. These
units may also be more desirable targets if the scientists are primarily
concerned with estimating the contemporary slip rate on a fault with a slip
rate that may vary over Quaternary timescales
\citep[e.g.,][]{rittase_temporal_2014,zinke_highly_2018}. For slow-moving 
faults, the slip either long waiting to be released, or recently released, may 
represent a sizeable fraction of the measured fault offset.

@@

The most commonly used is the \emph{exponential} distribution. This is
associated with a Poisson process, and is the distribution that results
from earthquakes being distributed uniformly randomly within some time interval.
Consequently, the probability of an earthquake (or other event) occurring at any
time does not change with time since the previous event (in other words, the
hazard function is time-invariant); this leads to the characterization of the
exponential recurrence distribution as `random', `memoryless' or
`time-independent'. The exponential distribution is also the simplest to
describe statistically, as it requires only one parameter (the mean rate
parameter), which is the inverse of the statistical scale parameter. (The scale 
parameter of a distribution determines the dispersion of the distribution, 
while the shape parameter determines the shape or form of the distribution; a 
parameter that shifts the distribution along the $x$-axis is called a location 
parameter.) The standard deviation of a large number of samples generated from 
an exponential distribution is equal to the mean.

@@

Clustered earthquake sequences have sets of very tightly spaced earthquakes
that are widely separated (Figure \ref{eq_spacing}, and have a \(CV>1\). These
may be generated from a hyperexponential distribution, which is the sum of
multiple exponentials with different means, or from the time-dependent
distributions given above, given the right parameters.

No consensus exists among earthquake scientists as to the most
appropriate recurrence interval distribution. As is generally the case
with propriety, the safest and probably most correct assumption is that
it is context-dependent. Many studies of plate boundary faults such as
the San Andreas conclude that major or `characteristic' earthquakes are
periodic\\
\citep[e.g.,][]{berryman_major_2012, scharer_quasi-periodic_2010}.
Conversely, many intraplate faults with low slip rates appear to show
clustered earthquakes separated by long intervals of seismic quiescence
\citep[e.g.,][]{clark_long-term_2012}. However, one can find examples of
studies indicating the opposite conclusions, even from the same study areas
\citep{tuttle_earthquake_2002, grant_paleoseismic_1994}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  A \emph{periodic} distribution, represented by a lognormal
  distribution with a mean recurrence interval \(\mu\) = 1000 years, and
  a standard deviation \(\sigma\) = 500 years, and a \(CV\) =
  0.5 \footnote{Please note that $\mu$ and $\sigma$ do not represent the scale
  and shape parameters of the lognormal distribution in this work, though these
  symbols are commonly used to represent these parameters elsewhere as the
  scale and shape parametersof a lognormal distribution $L$ are the standard
  deviation and mean of the normal distribution $\ln L$. Instead, the shape
  parameter $sh$
  and scale parameter $sc$ are derived from $\mu$ and $\sigma$ through the
  relations $sh = \sqrt{1 + \sigma^2 / \mu^2}$ and 
  $sc = \ln [ \mu / \sqrt{1 + \sigma^2 / \mu^2}]$.}
\item
  An \emph{unclustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 1000 years, and a \(CV\) =
  1.0.
\item
  An \emph{clustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 2000 years, and a \(CV\) =
  2.0.
\item
  An \emph{unclustered} time-independent distribution, represented by an
  exponential distribution with a mean recurrence interval \(\mu\) =
  1000 years, a standard deviation \(\sigma\) = 1000 years, and a
  \(CV\) = 1.0.
\end{enumerate}

@@

All earthquake recurrence distributions share a single earthquake slip
distribution (Figure \ref{slip_dist}). This distribution is a lognormal
distribution with \(\mu\) = 1 m and \(\sigma\) = 0.75 m, which produces
essentially `characteristic' earthquakes that still nonetheless have
some variability. This is representative of behavior observed in many
studies \citep[e.g.,][]{zielke_slip_2010, klinger_characteristic_2011,
zielke_earthquake_2018}. Taken together, the mean slip of 1 m and the
mean recurrence interval of 1,000 years shared by each of the recurrence
interval distributions yields a mean slip rate of 1 mm/yr. This rate is
fairly typical for intraplate faults studied by paleoseismologists, and
also allows for easy normalization so that the results of this study can
be generalized to faults with different parameters.

@@

The construction of the displacement histories is straightforward. From
each recurrence distribution, a little over 2,000 samples are drawn
randomly. Then, these are combined with an equal number of displacement
samples drawn randomly from the earthquake slip distribution. Finally, a
cumulative displacement history is created for each series from a
cumulative sum of both the recurrence interval samples (producing an
earthquake time series) and displacement samples (producing a cumulative
slip history). Years with no earthquakes are represented as having no
increase in cumulative displacement. Then, the series is trimmed at
year=2,000,000; it is initially made longer because the stochastic
nature of the sample sets means that 2,000 earthquakes may not always
reach 2,000,000 years.

@@

Please note that in the construction of the cumulative displacement
histories, all samples are independent. This means that the duration of any
recurrence interval does not depend on the duration of the previous or
subsequent interval (in other words, there is no autocorrelation in these
series); the same applies to the displacement samples. It is currently unknown 
to what degree autocorrelation exists in
real earthquake time and displacement series, or how much correlation is
present between recurrence intervals and subsequent displacements.
Autocorrelation in recurrence interval sequences is essentially unstudied,
though on the basis of preliminary, unreviewed analysis
\citep{styron_survival_2017}, I suspect that it is as important as \(CV\).

Furthermore, the magnitude of displacement is independent of the corresponding
recurrence interval. The framework of elastic rebound theory in its most basic
form should predict some correspondence between inter-event (loading) duration
and slip magnitude, and this is included (implicitly or explicitly) in
oscillator models incorporating complete stress or strain release in each
earthquake \citep[e.g.,][]{matthews_brownian_2002,dicaprio_post-seismic_2008}
or in any model where coseismic friction drops to zero, as this is functionally
equivalent (because \(f^a = \tau^a_s / \tau^a_n\), where \(f^a\), \(\tau^a_s\)
and \(\tau^a_n\) are respectively friction at rupture arrest, shear stress at
rupture arrest, and effective normal stress at rupture arrest, zero friction
implies zero shear stress, or complete stress drop). Given a reasonably
constant loading rate, complete shear stress or strain release implies some
proportionality between the loading time and displacement. Nonetheless, this
correspondence is not found in the more extensive paleoseismic datasets, such
as those by \citet{benedetti_earthquake_2013} (or the correlation may be
negative as found by \citet{weldon_wrightwood_2004}), but the number of
paleoseismic datasets of sufficient size and cquality to identify these effects
with statistical significance are few indeed.

@@

A major goal of this study is to provide an answer to the question, \emph{How
long should slip rates be measured over in order to estimate a meaningful
rate?} This question will be answered by looking at the distribution in
\(\hat{R}\) as a function of \(t\). Fifty values of \(t\) from 500 years to
100,000 years, logarithmically spaced, are used. Note that given \(\mu\) of
1000 years, this translates to 0.5--100 mean numbers of earthquakes in the
window.

@@

Note that with a measurement time exceeding 5--10 mean earthquake cycles, the
standard error ($n / \sqrt{(\sigma_D^2 + \sigma_r^2)}$, where $\sigma_D$ is
the standard deviation of displacement, and $\sigma_r$ is the standard
deviation of the recurrence interval) is a reasonable approximation for the
standard deviation of the variability in the slip rates due to
earthquake-cycle stochasticity, and shows broadly similar decreasing
variation with increasing earthquake cycles. However, the standard error is
symmetrical, though the variability displayed here is asymmetrical due to the
asymmetry of the recurrence interval and displacement distributions.

@@

Second, the distributions are asymmetrical, especially the 5--95\%
interval. The 95th percentile is generally several times as far from the
measured value as the 5th percentile, meaning that the true value of the
slip rate may be much greater than the measured rate but not a
commensurately small fraction of the measured rate.

Third, the median rate before convergence at 5 \(\mu\) is greater than
the true rate, meaning that in most cases, very short-term slip rate
measurements will underestimate the true slip rate; this is a systematic
bias. This is a particular concern for paleoseismological slip rate
estimates, where there are rarely more than 5 events in any given
trench, though this bias will be decreased if slip rate measurements are made
from closed intervals only. However, the median is less than 2 times
\(\hat{R}\) following just 2--3 events, so the systematic bias is unlikely to
be much greater than the measurement uncertainties on the age or offset of
the events.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The variability in slip rates calculated over time windows less than 5
  mean earthquake cycles is very large, but begins to stabilize
  following \textasciitilde{}10--20 earthquakes.
\item
  The most important factor in controlling the variability in slip rate
  estimates is the coefficient of variation (\(CV\)); the different
  distributions themselves are relatively unimportant. Faults with
  \emph{periodic} earthquakes, with \(CV < 1\), have less initial
  variability, and stabilize rapidly. Faults with \emph{unclustered}
  earthquakes, with \(CV = 1\), have more variability and stabilize
  more slowly. Faults with \emph{clustered} earthquakes,\\
  with \(CV > 1\), have a great amount of initial variability and
  require a large number of earthquakes to stabilize.
\item
  The epistemic uncertainties around a measured slip rate are similarly
  large initially and then decrease with time. These uncertainties are
  initially biased, such that the measured slip rates typically
  underestimate the true slip rate for \textless{}5 mean earthquake
  cycles, but this fades with time. However, the uncertainties remain
  asymmetric, with a strong right skew.
\end{enumerate}
